Lipid discovery enabled by sequence statistics and machine learning

Version 1

Abstract

Abstract

Bacterial membranes are complex and dynamic, arising from an array of evolutionary pressures. One enzyme that alters membrane compositions through covalent lipid modification is MprF. We recently identified that Streptococcus agalactiae MprF synthesizes lysyl-phosphatidylglycerol (Lys-PG) from anionic PG, and a novel cationic lipid, lysyl-glucosyl-diacylglycerol (Lys-Glc-DAG), from neutral glycolipid Glc-DAG. This unexpected result prompted us to investigate whether Lys-Glc-DAG occurs in other MprF-containing bacteria, and whether other novel MprF products exist. Here, we studied protein sequence features determining MprF substrate specificity. First, pairwise analyses identified several streptococcal MprFs synthesizing Lys-Glc-DAG. Second, a restricted Boltzmann machine-guided approach led us to discover an entirely new substrate for MprF in Enterococcus, diglucosyl-diacylglycerol (Glc2-DAG), and an expanded set of organisms that modify glycolipid substrates using MprF. Overall, we combined the wealth of available sequence data with machine learning to model evolutionary constraints on MprF sequences across the bacterial domain, thereby identifying a novel cationic lipid.

Introduction

Lipids are ubiquitous and diverse group of hydrophobic and amphiphathic compounds that are critical for fundamental biological processes, including the formation of cell membranes, protection, and cargo delivery; energy storage; and cell signaling pathways [1]. They are relatively smaller than other complex biomolecules, such as proteins, thereby allowing a larger portion of their surface to interact with other macromolecules. Chemical modifications to lipids, including changes in fatty acid tails (saturated and unsaturated) [2], modifications of head groups [3], and alterations to lipid charge [4], impact these interactions, as well as the physicochemical properties of membranes. Commonly, bacteria synthesize lipids that are negatively charged or neutral [5]. Some bacteria modify the negatively charged lipids with amino acids to make them positively charged to confer resistance to cationic antimicrobial peptides and cationic antibiotics [6, 7]. The bacterial domain is a rich source of natural lipids that are likely co-evolved for specific interactions with host tissues and other cells, yet the full extent of this chemical diversity remains unexplored. Finding bacteria that possess unique lipids can be useful for biotechnology. For example, natural lipids found in bacteria can be used for drug delivery. A study by [8] used low cytotoxic outer membrane vesicles (OMV) that contained a modified lipopolysaccharide (LPS) to deliver siRNA to targeted cancer cells. This delivery mechanism had success and demonstrates the possibility of using naturally occurring bacterial lipids for future drug delivery processes to eukaryotic cells.

One enzyme responsible for modifying lipids in bacteria is the multiple peptide resistance factor (MprF). MprF modifies lipids through the transfer of amino acids from charged tRNAs to the head group of the anionic membrane phospholipid, phosphatidylglycerol (PG) [4]. The MprF protein is comprised of two domains: an N-terminal flippase domain responsible for moving aminoacylated lipids from the inner membrane leaflet to the outer leaflet and a C-terminal synthase domain responsible for the aminoacyl transfer from tRNA to lipids [9, 10]. Recently, in Streptococcus agalactiae (Group B Streptococcus, GBS) it was found that MprF can modify two different substrates with lysine (Lys) – the glycolipid glucosyl-diacylglycerol (Glc-DAG) and the phospholipid, PG [11], generating Lys-Glc-DAG, as well as Lys-PG [11, 12] (Figure 1). An S. agalactiae mprF deletion mutant no longer synthesizes Lys-Glc-DAG or Lys-PG and expression of S. agalactiae mprF in the heterologous host Streptococcus mitis conferred Lys-Glc-DAG and Lys-PG synthesis to S. mitis [11]. This was the first time MprF was demonstrated to add lysine onto a neutral glycolipid (Glc-DAG).

Chemical structures of lysine lipids. Lys-PG, lysyl-phosphatidylglycerol; Lys-Glc-DAG, lysylglucosyl-diacylglycerol; Lys-Glc2-DAG, lysyl-diglucosyl-diacylglycerol.

The synthesis of Lys-Glc-DAG by S. agalactiae MprF illustrated a unique lipid substrate and product by the enzyme that had not been characterized before, highlighting the possibility for further exploration into unknown lipids and lipid substrates by MprF of different bacterial species. We sought to investigate the molecular determinants of enzyme specificity and identify other bacterial species that may have this novel specificity. This investigation led us from the standard methodology of BLASTp queries [13], which are based on single residue amino acid substitution frequencies, to a more modern statistical method called a Restricted Boltzmann Machine (RBM) [14], which captures global statistical patterns within natural sequence data. In this work, we develop a general strategy for combining this high-powered statistical model with lipidomic analysis to discover MprF variants with unique enzymatic activity. This process led us to the discovery of an uncharacterized lipid substrate for MprF, diglucosyl-diacylglycerol (Glc2-DAG), which is present in several strains identified by the parameters of the statistical model. These models also provide interpretable and actionable information for use in further enzyme characterization, and we show that this application is a useful companion tool for experimental work.

Results

Simple pairwise sequence statistics identify streptococcal MprF enzymes that synthesize Lys-Glc-DAG

Utilizing a method based on the amino acid sequence of S. agalactiae MprF, we identified four streptococcal MprFs with high sequence identity to S. agalactiae COH1 MprF (WP 000733236.1). A BLASTp analysis found that Streptococcus sobrinus ATCC 27352 MprF (WP 019790557.1) had 61.4% amino acid identity to the S. agalactiae MprF, Streptococcus salivarius ATCC 7073 MprF (WP 002888893.1) had 43.09% amino acid identity, Streptococcus ferus MprF (WP 018030543.1) had 61.88% amino acid identity, and Streptococcus downei (WP 002997695.1) had 61.03% amino acid identity (Table 1). Plasmids were designed to express the S. sobrinus mprF (pSobrinus), S. salivarius mprF (pSalivarius), S. ferus mprF (pFerus), and S. downei mprF (pDownei) genes. The plasmids were transformed into the expression host S. mitis NCTC12261 (SM61). The empty vector (pABG5Δ phoZ, referred to as pABG5 here) was also transformed (Figure 2a). S. mitis does not natively encode mprF but synthesizes lipids (Glc-DAG; PG) that are substrates for S. agalactiae MprF [15, 16, 11], making it an appropriate heterologous host for expressing streptococcal MprFs. Previously, a plasmid expressing S. agalactiae mprF (pGBSMprF) [11] was generated and transformed into S. mitis. Lys-Glc-DAG and Lys-PG presence in SM61 was only possible with the expression of pGBSMprF (Figure2b) [11].

Synthesis of lysine lipids (Lys-PG and Lys-Glc-DAG) in S. mitis expressing mprFs from S. agalactiae, S. salivarus, and S. ferus. (a) S. mitis NCTC12261 with empty vector control (pABG5) lacks lysine lipids; (b) S. agalactiae mprF (pGBSMprF) produces both Lys-PG and Lys-Glc-DAG; (c) S. salivarius mprF produces only Lys-PG; (d) S. ferus mprF produces only Lys-Glc-DAG. Left panels: total ion chromatograms (TIC); middle panels: mass spectra of retention time 19.5–21.5 min showing Lys-PG and PC; right panels: mass spectra of retention time 26–30 min showing Lys-Glc-DAG. Note: “*” is an extraction artifact due to chloroform used. DAG, diacylglycerol; MHDAG, monohexosyldiacylglycerol; DHDAG, dihexosyldiacylglycerol; PG, phosphatidylglycerol; Lys-PG, lysyl-phosphatidylglycerol; Lys-Glc-DAG, lysyl-glucosyl-diacylglycerol; PC, phosphatidylcholine.

S. mitis lipids of strains expressing the plasmids pSobrinus (Table 1), pSalivarius (Figure 2c), pFerus (Figure 2d) and pDownei (Table 1) were analyzed. Lipidomic analysis found that pSobrinus, pDownei, and pFerus conferred synthesis of Lys-Glc-DAG, but not Lys-PG in S. mitis. Notably, pFerus confers synthesis of Lys-Glc-DAG at a similar level to S. agalactiae MprF (Figures 2b, d). In contrast, pSalivarius conferred synthesis of Lys-PG only and not Lys-Glc-DAG. Importantly, S. salivarius MprF also synthesized Lys-PG at a level most similar to S. agalactiae MprF (Figures 2b, c). Lysine-modified lipids synthesized by S. mitis strains expressing streptococcal MprFs used in this study are summarized in Table 1. In general, streptococcal MprF enzymes with higher amino acid percent (> 60%) to S. agalactiae MprF conferred Lys-Glc-DAG synthesis to S. mitis, but only S. agalactiae MprF conferred both Lys-Glc-DAG and Lys-PG synthesis in S. mitis.

Restricted Boltzmann Machines can provide sensitive, rational guidance for sequence classification

We set out to understand which specific MprF amino acid residues are involved in enzyme specificity for either the PG or Glc-DAG substrate. To this end, we employed a Restricted Boltzmann Machine (RBM), a probabilistic graphical model which aims to model the probability of data in a data set through statistical connections between the features of the data and a set of hidden units (weights) [17]. This class of models is closely related to another model, the Boltzmann Machine, which has been very successful in elucidating important residues in protein structure and function [18], one example being the careful tuning of enzyme specificity in bacterial two-component regulatory systems [19, 20]. The Boltzmann Machine formulation from [18], termed Direct-Coupling Analysis (DCA), stores patterns in the form of a pairwise coupling matrix and local field matrix (shown in Equation 2), which has been useful in understanding how likely pairs of residues are to interact in physical space. We hypothesize that S. agalactiae MprF and other MprF enzymes that utilize glycolipid substrates have unique patterns of interacting residues relative to those enzymes that use only PG as a substrate.

While modeling how pairs of residues covary is powerful, many important features of enzyme function like allostery or enzyme specificity (which is the purpose of this analysis) may involve more than two residues [21], and in these situations it can be difficult to connect sets of residues using solely pairwise parameters [22]. Therefore, a more global approach is required to address our hypothesis. To this end, a RBM has recently been developed for use in the learning of protein sequence data, with special attention given to producing a model which produces sparse, interpretable, and biologically meaningful representations of these higher-order statistical couplings within the hidden units [14]. The design of this RBM (see the marginal distribution form shown in Equation 3) allows the following useful properties: higher-order couplings between residues (sets of coupled residues instead of pairs); bimodal hidden unit outputs given the training data set (Equation 4); sparse, interpretable configurations of the weights (wμ) (Equation 5) and a compositional representation of weights [23], where input sequences are modeled largely through combinations of hidden units instead of single, highly activating units.

The compositional representation is a critical feature of this method, allowing us to find independently coupled sets of residues which in combination describe these protein sequences, and it is achieved through the particular choices of training parameters. More details on the training parameter choices are found in Methods and Materials. In particular, we focus on the scoring of sequences using individual weights wμ: where v is an input sequence vectorized through one-hot encoding, to produce an output score (a single number) which depends on the input sequence residues at position i and the value of the weight matrix for that residue at that position.

We used this methodology to study the sequences in the family of MprF, which is a large protein composed of independent flippase and synthase domains [9, 24]. In our analysis, we restrict ourselves to the cytosolic region (see yellow box in Figure 3a), also known as the aminoacyl-PG synthase domain, which constituted the Pfam family DUF2156 (now renamed LPG synthase C) [25]. One weight in particular is highlighted, where we find bimodal activation (Figure 3b) of the hidden unit (Figure 3c) when Equation 1 is evaluated with the training sequences. Interpreting the sequence histograms, a positive value output in Figure 3b corresponds to a sequence containing combinations of the residues listed in the positive portion of the weight diagram in Figure 3c, whereas a negative output corresponds to residues in the negative portion. For display purposes, the residues shown in all of our weights correspond to the positions where the magnitude of the visible positions wiμ are greater than some threshold proportional to the largest magnitude position (position 212); the weight has values for all possible amino acids in the entire protein length, but they are considerably smaller than this threshold and generally very low.

Example of hidden unit analysis and usage. (a) The structure of PDB:7DUW, with the red colored region being the transmembrane flippase domain and the yellow boxed region the cytosolic domain which we focus on. (b) The scores produce by inputting a sequence into a hidden unit, producing a single number as output which corresponds to a summation of negatively and positively weighted residues. Performed on entire training set (histogram in blue), highlighting sequences corresponding to predominantly positive weighted residues. (c) Hidden unit from an RBM trained on the Pfam DUF2156 domain. The MSA positions 152 and 212 correspond to residues S684 and R742, respectively. (d) Residues (in yellow) in the MprF cytosolic domain which form the binding pocket for Lys-tRNALys (the ligand analogue L-lysine amide shown in green), from PDB:4V36. LYN, L-lysine amide.

Two positions in particular, 152 and 212, in the positive portion of the weight correspond to residues 684 and 742, which were experimentally demonstrated to be relevant for aminoacylated-PG production by the Bacillus licheniformis and Pseudomonas aeruginosa cytosolic synthase domain variants [26]; in B. licheniformis the residues which form a complex with the lysine from Lys-tRNALys are shown in Figure 3c. We have highlighted these two aforementioned variants as well as the S. agalactiae variant for demonstration of sequence scoring through Equation 1 in Figure 3a. We used this weight to create a filtered data set; only sequences that contained this necessary motif were of interest to us, and we subsequently trained a new model using only sequences with a catalytic weight output greater than 2. We outline this general workflow in Figure 4.

Schematic of the RBM methodology. An aligned set of protein sequences is first used to learn a hidden unit representation that best describes the statistics of the sequence dataset given restrictions on the hidden unit representation. Then, the individual hidden units can be studied to find particular units which allow useful enzyme classification, and additionally, these weights can be meaningfully interpreted as statistically co-varying sequence configurations. Additionally, the classification can be used to create filtered datasets to train more models.

RBM weights identify plausible sequence residues for functional characterization

After filtering the data set and training another model, we set out to find weights which could describe MprF specificity. To this end, we identify weights which are sparse, have high overall magnitude, and involve residues which could be plausibly linked to the lipid binding specificity (see Methods for more details). Through this search method we identified the two hidden units shown in Figure 5a. They both have high sparsity and high magnitude relative to the majority of other weights in the model, and importantly they correspond to residues localized to regions of the catalytic domain that had been previously identified through Autodock experiments using a PG molecule with side chains C5:0/C8:0 [26] (highlighted in Figure 5a). The scoring of the training set of sequences with these weights is shown in Figure 5b, and we see a clear separation between the cytosolic domain sequences which produces Lys-Glc-DAG (from S. agalactiae) and two domain variants which produce only aminoacylated-PG (from P. aeruginosa and B. licheniformis). Importantly, these output scores depend on a greatly reduced subset of the full sequence and do not correspond to clustering on total sequence identity (Figure S1).

Proposed set of weights for classifying lipid specificity. (a) two hidden units found in an RBM trained on the filtered Pfam dataset. The hidden unit residues are highlighted in the PDB:4V34 structure, with arrows pointing to their corresponding residue sets. (b) the outputs of the weights when scoring the sequences used in the training set and sequences from NCBI not used during training (N=23,138). S. agalactiae produces Lys-Glc-DAG and Lys-PG, while B. licheniformis and P. aeruginosa produces only aminoacylated-PG. Q1-Q4 are quadrant labels which we refer to throughout the paper.

Using these two weights as sequence classifiers with a potential link to the substrate specificity of cytosolic domain variants, we used Figure 5b to propose sequences in quadrant three (Q3), the quadrant occupied by the S. agalactiae MprF variant with this novel glycolipid-modifying function. We selected a sequence which was distant in terms of sequence identity from the Streptococcus genus (Table S1) and which has been implicated in human pathologies [27], Enterococcus dispar MprF.

Enterococcal MprF enzymes synthesize Lys-Glc-DAG, Lys-PG, and a novel cationic glycolipid Lys-Glc2-DAG

Enterococcus dispar ATCC 51266 was identified as a possible candidate for novel lipid synthesis through the analyses described above. Remarkably, our lipidomic analysis found that E. dispar synthesizes a novel cationic lipid, lysyl-diglucosyl(Glc2)-diacylglycerol (Lys-Glc2-DAG) (Figure 1, Figure 6a,b), as well as Lys-Glc-DAG and Lys-PG (Figure 6c, Figure S2). The structural identification of Lys-Glc-2-DAG was supported by exact mass measurement ([M+H]+ observed at m/z 1047.730; calculated m/z 1047.731) and tandem MS (Figure 6b). Lys-Glc-2-DAG is unusually polar and charged for a lipid; chromatographically, Lys-Glc2-DAG is highly retentive on a silica-based normal phase column and elutes off the column at the very end of the gradient (Figure S2). The use of an amino-based HPLC column led to much improved elution profiles for Lys-Glc2-DAG and other glyco- and lysine lipids (Figure 6)

Identification of Lys-Glc2-DAG in E. dispar. (a) Positive ion mass spectrum of Lys-Glc2-DAG species in E. dispar. (b) MS/MS product ions and fragmentation scheme of Lys-Glc2-DAG. (c) Extracted ion chromatograms of LC/MS of lysine lipids (Lys-PG, Lys-Glc-DAG, Lys-Glc2-DAG) and Glc2-DAG separated on an amino HPLC column. Glc2-DAG, diglucosyl-diacylglycerol; Lys-PG, lysyl-phosphatidylglycerol; Lys-Glc-DAG, lysyl-glucosyl-diacylgylcerol; Lys-Glc2-DAG, lysyl-diglucosyl-diacylglycerol.

The discovery of Lys-Glc2-DAG in Enterococcus dispar led us to analyze other Enterococcus strains: E. faecium 1,231,410, E. faecalis T11 and OG1RF E. gallinarium EG2, E. casseliflavus EC10, and E. raffinosus Er676. All Enterococcus strains examined were found to synthesize Lys-PG, Lys-Glc-DAG and Lys-Glc2-DAG at varying levels (Figure 7a and Table 2). MprF-dependent Lys-PG production by E.faecalis and E. faecium was previously reported [28, 29], but cationic glycolipids production has not been previously reported. E. gallinarium EG2 and E. casseliflavus EC10 encode a single copy of mprF gene, while the other enterococci we tested encode two distinct mprF genes (mprF 1 and mprF 2) [28, 29]. When these MprF sequence variants are plotted using our chosen RBM weights, we see that the two single copy strains plot in quadrant two (Q2), while for the two copy sequences, each MprF variant maps to separate coordinates (Figure S3). Both E. raffinosus Er676 MprF variants plot to quadrant two (with the lowest Weight 2 value being 0.34). The E. faecium and E. faecalis variants have copies plotting directly in quadrant one and variants plotting farther out from the sequence cluster; we note that the two displaced variants are both lacking the characteristic aspartic acid and glutamic acid amino acids at positions 100 and 48 in the weights (Figure 5a), which are the dominant contributions for determining quadrant one occupancy. To clarify which variants were determinants of the novel cationic lipid production, lipidomic analysis on an E. faecalis OG1RF mini-mariner (EfaMarTn) transposon mutant [30] with a transposon insertion within OG1RF 10760 mprF (referred to as mprF 2 in previous studies [28]; OG1RF 10760::Tn) revealed that Lys-PG, Lys-Glc-DAG and the newly identified Lys-Glc2-DAG were absent when the synthase domain of the gene was disrupted (Figure 7b, Table S1). This led us to the conclusion that E. faecalis mprF 2 is necessary for the synthesis of the three Lys-lipids in E. faecalis. When the empty vector (pABG5) was transformed into OG1RF 10760::Tn, no Lys-lipids synthesis was observed as expected (Figure 7c). This phenotype was complemented by expression of E. faecalis mprF 2 from a plasmid (pOGMprF2) (Figure 7d). Additionally, expression of E. faecium mprF1 (EFTG 00601) (pEfMprF1) restored Lys-PG and Lys-Glc2-DAG synthesis to the E. faecalis OG1RF 10760::Tn strain (Figure 7e), while E. faecium mprF2 (EFTG 02430) did not rescue Lys-lipid synthesis (Figure 7f). Taken together, these data indicate that E. faecalis MprF2 (OG1RF 10760) and E. faecium MprF1 (EFTG 00601), like S. agalactiae MprF, act on both phospholipid and glycolipids substrates, additionally synthesizing the novel cationic glycolipid, Lys-Glc2-DAG.

E. faecalis MprF2 and E. faecium MprF1 confer Lys-Glc2-DAG synthesis. (a) OG1RF-WT; (b) OG1RF 10760::Tn lacks lysine lipids; (c) OG1RF 10760::Tn + pABG5 lacks lysine lipids; (d) OG1RF 10760::Tn + pOGMprF2 restores lysine lipids; (e) OG1RF 10760::Tn + pEFMprF1 restores lysine lipids; (f) OG1RF 10760::Tn + pEFMprF2 lacks lysine lipids. Expression of OGMprF2 and EFMprF1 in OG1RF 10760 Tn mutant restore Lys-Glc2-DAG synthesis. Shown are the extracted ion chromatograms of lysine lipids (Lys-PG, Lys-Glc2-DAG) and Glc2-DAG separated on an amino HPLC column. Note: Lys-Glc-DAG was found in trace amounts or missing from lipid extractions. Glc2-DAG, diglucosyl-diacylglycerol; Lys-PG, lysyl-phosphatidylglycerol; Lys-Glc2-DAG, lysyl-diglucosyl-diacylglycerol.

Weight 2 is correlated with GlcN-DAG substrate activity in MprF

When we started our RBM analysis, we had substrate specificity information for only the quadrants one and three, and through identifying the novel substrate in E. dispar we expanded our testing to quadrant two by testing Enterococcus strains. To expand our RBM analysis, we tested specific sequences which were found in quadrant four (Q4). We conducted lipid analysis on Ligilactobacillus salivarius ATCC 11741, Levilactobacillus brevis ATCC 14869, Lacticaseibacillus rhamnosus ATCC 7469, Lactobacillus casei ATCC 393, and Lacticaseibacillus paracasei ATCC 25302. All strains tested synthesize Lys-PG (Table 2). Only Levilactobacillus brevis and Lacticaseibacillus paracasei synthesized Lys-Glc2-DAG, L. paracasei synthesized low levels of Lys-Glc2-DAG. An additional quadrant one (Q1) strain Exiguobacterium acetylicum UTDF19-27C, was analyzed by lipidomics. It was found to synthesize only Lys-PG. This MprF falls near Staphylococcus aureus RN4220 and Bacillus subtilis 168 in our plot, both of which synthesize only Lys-PG [31, 24] (and experimentally confirmed in this study). This further supports that strains located in quadrant one of Figure 8 are capable of Lys-PG synthesis.

The two proposed hidden unit outputs with all of the sequences from Table 2 labeled. Protein ID of highlighted sequences are listed in Table S1. (#) indicates the lipid activity was confirmed through heterologous expression.

Summaries of all species/variants tested and the Lys-lipids they produced are shown in Table 2, and plotted in the two RBM weights in Figure 8. We list for each MprF variant which of the three tested lipids it acts on, and in Figure 8 we indicate whether a variant has only PG as a substrate or if it operates on GlcN-DAG as a substrate irrespective of its PG activity. With this plot, we can see that sequences which plot with positive values in Weight 2 are more likely to have GlcN-DAG as a substrate (8/9), and sequences with negative values predominantly use PG (7/11), with a Fisher’s exact test p-value of 0.028 (Table S2). Weight 1, however, is not correlated with GlcN-DAG specificity, with a Fisher’s exact p-value of 0.67. Therefore we conclude that Weight 2, and thus the high magnitude sequence positions identified by it, are correlated with GlcN-DAG specificity.

Discussion

The utilization of the S. agalactiae MprF amino acid sequence as a query for BLASTp was a simple method to identify various streptococcal MprFs and further our understanding of MprF specificity. This allowed us to identify three MprFs that can synthesize the highly cationic Lys-Glc-DAG. Importantly, although highly similar to S. agalactiae MprF, this method did not identify other streptococcal MprFs capable of lysine addition to both glycolipid and phospholipid substrates (generating Lys-Glc-DAG and Lys-PG) nor did it lead to the identification of a novel cationic glycolipid found in Enterococcus. Sequence logos [32] may identify important sequence features, but largely in instances where amino acids are highly conserved. When amino acids must interact to produce a specific function, yet the exact identity of those amino acids is not a requirement, methods which model the strength of amino acid coevolution in sequence sets can find important signals which in the sequence logo might appear to be negligible. For this reason we used the Restricted Boltzmann Machines (RBM) method [14], which can model statistical connectivity of residues at multiple sites in the protein sequence.

The RBM can be used in a number of ways to study sequence data. Firstly, it allowed us to filter the sequence data using hidden units (Figure 3); the two weights which formed the basis of our specificity analysis were not found in models trained on the Pfam dataset when it was initially pulled from Pfam, and this was potentially due to the presence of sequences without relevant catalytic function adding noise to the dataset. The most prominent feature for our purposes is the ability to use the hidden units learned through RBM training to classify sequences. The bimodal nature of this particular model provides a relatively simple interpretation which is straightforward to use for prediction. We show in Figure 5 that the combination of clustering and interpretable weight structure allows us to identify a small subset of residues within a structure, grounding our statistical clustering in MprF’s structural features. The combination of experimental evidence and evolutionary information provided a strong rationale for selecting organisms for further study, ultimately leading to the discovery of novel cationic lipid biosynthesis in Enterococcus dispar (Figure 6, Figure S2).

One goal of this study was to find the structural determinants of GlcN-DAG activity by MprF, and we found that Weight 2 from our RBM analysis was predictive of this. However, there are notable exceptions such as the E. faecium allele EGTF 00601, which has sequence features placing it in quadrant one yet has GlcN-DAG activity. Therefore, identifying the exact sequence determinants of GlcN-DAG specificity remains to be fully elucidated. One explanation could be our exclusion of the N-terminal flippase domain and the non-cytosolic region of the C-terminal domain in the RBM analysis, where important residue interactions determining specificity could occur and is under further investigation. Additionally, a challenge in experimental validation will be to standardize the method of studying mprF alleles from diverse organisms. When using native organisms, lipids produced may vary during the growth phase of the organisms, the media they are grown in, and whether a membrane stressor is present (i.e. inducible production of specific lipids). For this study we used stationary phase cultures of equal volume and bacteria were grown in their respective standard laboratory media. The use of a heterologous host for expression of mprF alleles is a method of standardization, but may result in not identifying novel lipids that would be found in their native bacteria since the appropriate lipid substrates may not be present [16, 12]. A combination of both approaches (native organisms and heterologous hosts) may be required. Ultimately, mutational studies of the residues identified by the RBM will allow for identification of the residues critical for specificity, though likely the residues will depend in part on the primary sequence being altered.

The general method of RBM-guided exploration has applications in enzyme design, for example, in enhancing methods like directed evolution [33]. Concretely, the residues identified through weights can be thought of as templates for mutational exploration, greatly reducing the search space for assessing which residues are critical for specificity determination. Enzymes like MprF that produce lipids of modified charge have potential applications in biotechnology and therapeutics, where the design of lipid nanoparticles with different charge and chemical properties of the head group have important physiological implications [34]. Notably, the identification of Lys-Glc2-DAG opens new research avenues, particularly in the fields of antibiotic resistance and host-pathogen interactions, wherein lipid modification and cationic lipids play key roles. With respect to human health, it would be relevant to investigate whether Lys-Glc2-DAG plays a role in enterococcal resistance to the last-line membrane-active antibiotic, daptomycin.

Methods and Materials

Bacterial Strains and Growth Conditions

See Tables S1 and S3 for a full list of bacterial strains used in this study. Streptococcus mitis NCTC12261 (referred to as SM61 here) [35] was grown in Todd Hewitt Broth (THB) at 37°C with 5% CO2. Escherichia coli DH5α [36] and Bacillus subtilis 168 [37] was grown at 37°C with shaking at 225 rpm in lysogeny broth (LB). Staphylococcus aureus RN4220 [38] was grown at 37°C with shaking at 220 rpm in Trypic Soy Broth (TSB). Streptococcus agalactiae COH1 ATCC BAA-1176 [39] and CJB111 [40, 41] were grown in THB at 37 °C. Enterococcus faecium 1,231,410 [42], Enterococcus faecalis T11 [43] and OG1RF [44], Enterococcus gallinarium EG2 [42], Enterococcus casseliflavus EC10 [42], and Enterococcus raffinosus Er676 [45] were grown at 37°C in Brain Heart Infusion (BHI). Enterococcus dispar ATCC 51266 [46] was grown at 30°C in BHI. Exiguobacterium acetylicum UTDF19-27C was grown in BHI at 37 °C. Ligilactobacillus salivarius ATCC 11741 [47], Lacticaseibacillus rhamnosus ATCC 7469 [48], Lactobacillus casei ATCC 393 [49] and Lacticaseibacillus paracasei ATCC 25302 [48] were grown in Lactobacilli MRS Broth at 37°C and 5% CO2. Levilactobacillus brevis ATCC 14869 [50] was grown in Lactobacilli MRS Broth at 30°C, 5% CO2. Bacillus licheniformis ATCC 14580 was grown in nutrient broth at 37°C with shaking at 225 rpm [51]. A transposon mutant of E. faecalis OG1RF from [30] with a mini mariner transposon (EfaMarTn) insertion in OG1RF 10760 (OG1RF 10760::Tn) was grown on BHI supplemented with chloramphenicol at a concentration of 15 μ/mL. The mutant was confirmed to be erythromycin-sensitive. The transposon mutant location was confirmed through Sanger sequencing performed by the Genome Center at the University of Texas at Dallas (Richardson, TX).

Where appropriate for plasmid selection, kanamycin (Sigma-Aldrich) was added. For E. coli, a concentration of 50 μg/mL was used, for S. mitis a concentration of 300 μg/mL, and for E. faecalis a concentration of 500 μg/mL was used.

Routine Molecular Biology Procedures

Genomic DNA was extracted as done previously in [52] and [53]. All PCR reactions used Phusion polymerase (Thermo Fisher Scientific) and Phusion 5x HF buffer (Thermo Fisher Scientific) in a Veriti PCR machine (Applied Biosystems). List of primers used are found in Table S4. Gibson assemblies were completed using a 2x HI-FI Assembly master mix following the manufacturer’s protocol (New England Biolabs). PCR clean-up was done using the GeneJET PCR Purification Kit (Thermo Fisher Scientific) per manufacturer protocol. Plasmid extractions were performed per manufacturer protocol using the GeneJET Plasmid miniprep kit (Thermo Fisher Scientific). All plasmid constructs were confirmed by Sanger sequencing at the Massachusetts General Hospital CCIB DNA Core facility or by Illumina sequence at SeqCenter. DNA concentrations were measured using Nanodrop (Thermo Fisher Scientific) or Quibit 2.0 (Invitrogen by Life Technologies). Optical Density at 600 nm (OD600nm) was measured in a disposable cuvette (Thermo Fisher Scientific) using a spectrophotometer (Thermo Scientific Genesys 30).

Gibson assembly was performed using pABG5 as previously described [11]. Transformation of plasmids into E.coli, S. mitis, and E.faecalis was performed as previously described in [11, 16, 54, 55].

Acidic Bligh-Dyer Lipid Extractions

Bacteria were grown for approximately 16 hours overnight in 15 mL of an appropriate culture medium. After growth, OD600nm measurements were taken, and cells were pelleted at 4,280 x g for 5 minutes at room temperature in a Sorvall RC6+ centrifuge. The supernatant was tipped out, and the cell pellet was washed and resuspended in 1X Phosphate Buffered Saline (PBS). The cells were pelleted again, and all the supernatant was aspirated out. The cell pellet was stored at −80°C until lipid extraction. An acidic Bligh-Dyer lipid extraction was performed as previously reported [16, 12]. Briefly, cell pellets were resuspended in 0.8 mL of 1x Dulbecco’s PBS (Sigma-Aldrich). Cells were transferred to a 9 mL glass tube with a Teflon-lined cap (Pyrex). 1 mL chloroform (MilliporeSigma) and 2 mL methanol (Thermo Fisher Scientific) were added to create the single-phase Bligh-Dyer. Tubes were vortexed every 5 minutes for 20 minutes, then centrifuged at 500 x g for 10 minutes at room temperature. The supernatant was transferred to a new 9 mL glass tube. 100 μL of Hydrochloric acid (Thermo Fisher Scientific) was added followed by 1 mL of chloroform and 0.9 mL of 1x Dulbecco’s PBS to create the two-phase Bligh-Dyer. Tubes were gently mixed and centrifuged at 500 x g for 5 minutes at room temperature. After centrifuging, the bottom layer was extracted to a new tube and dried under nitrogen gas and stored at −80°C prior to lipidomic analysis. Lipid analyses were repeated in biological triplicate aside, from OG1RF 10760::Tn extractions, which were performed once.

Analysis of lysine lipids by an amino column-based Liquid Chromatography-Electrospray Ionization-Mass Spectrometry (LC-ESI MS): A new method

An amino column-based normal phase LC-ESI MS was performed using an Agilent 1200 Quaternary LC system coupled to a high-resolution TripleTOF5600 mass spectrometer (Sciex, Framingham, MA). A Unison UK-Amino column (3 μm, 25 cm × 2 mm) (Imtakt USA, Portland, OR) was used. Mobile phase A consisted of chloroform/methanol/aqueous ammonium hydroxide (800:195:5, v/v/v). Mobile phase B consisted of chloroform/methanol/water/ aqueous ammonium hydroxide (600:340:50:5, v/v/v/v). Mobile phase C consisted of chloroform/methanol/water/aqueous ammonium hydroxide (450:450:95:5, v/v/v/v). The elution program consisted of the following: 100% mobile phase A was held isocratically for 2 min and then linearly increased to 100% mobile phase B over 8 min and held at 100% B for 5 min. The LC gradient was then changed to 100% mobile phase C over 1 min and held at 100% C for 3 min, and finally returned to 100% A over 0.5 min and held at 100% A for 3 min. The total LC flow rate was 300 μl/min. The MS settings were as follows: Ion spray voltage (IS) = 5000 V, Curtain gas (CUR) = 20 psi, Ion source gas 1 (GS1) = 20 psi, De-clustering potential (DP) = 50 V, and Focusing Potential (FP) = 150 V. Nitrogen was used as the collision gas for MS/MS experiments. Data acquisition and analysis were performed using Analyst TF1.5 software (Sciex, Framingham, MA).

Generation of mprF expression plasmids

The S. downei mprF nucleotide sequence (Locus tag HMPREF9176 RS03810) and S. ferus mprF nucleotide sequence (Locus tag A3GY RS0106165) were used to design synthetic geneblocks (GeneWiz from Azenta Life Sciences). A 20 nucleotide 5’ extension and a 20 nucleotide 3’ extension complementary to the pABG5 plasmid were added to enable Gibson assembly of the inserts with pABG5. Sequences for the geneblocks can be found in Table S5. The S. salivarius mprF (Locus tag SSAL8618 04345) was amplified from Streptococcus salivarius ATCC 7073. A 20 nucleotide 5’ extension and a 20 nucleotide 3’ extension complementary to the pABG5 plasmid was added to the insert. The same protocol was used for generation of pSobrinus, pEFM-prF1, pEFMprF2, and pOGMprF2 with slight modifications. S. sobrinus mprF (Locus tag DLJ52 05040) was amplified from Streptococcus sobrinus ATCC 27352. E. faecium mprF1 (Locus tag EFTG 00601) and E. faecium mprF2 (Locus tag EFTG 02430) were amplified from Enterococcus faecium 1, 231, 410. E. faecalis mprF2 (Locus tag OG1RF 10760) was amplified from E. faecalis OG1RF.

Boltzmann Machines and Restricted Boltzmann Machines for Protein Sequences

Formulation of models

The Boltzmann Machine, as described in [18], is defined as: which is a probability distribution defined on the pairwise interactions (the couplings eij) between the positions (vn) in an input sequence (v) and a local field term hi. The pairwise coupling matrix is akin to a covariance matrix, and the local fields are similar to single site frequency measures.

The joint probability distribution of the Restricted Boltzmann Machine as described in [14] is as follows: Here, the vector v again represents the input sequence data, g(vi) is a local field which controls the conditional probability of the input data, 𝒰μ is the hidden unit potential/activation, and the terms hi and wiμ(v) couple the input variables with the hidden unit variables. In this way, interactions between residues are mediated by a relatively smaller set of hidden units which do not interact directly with each other, leading to a bipartite structure as opposed to the Boltzmann machine’s fully connected pair based formulation.

Particularly important is the potential function 𝒰, which in this work is the dRELU function defined as: The four parameters γμ,±, θμ,± are learned through the inference procedure, and allow the positive and negative components of the weights wμ to be separately gated, which allows learning of bimodal distributions for the output hμ. Another key feature is the sparsity regularization on the weights applied during model inference: where q is the number of amino acids plus a gap character (21) and the hyperparameter can be increased to induce sparsity and lower weight magnitude. Equations 4 and 5 together guide the representation of the hidden units to bimodal outputs with interpretable features.

Dataset acquisition and preprocessing

Multiple sequence alignment used for model training was the DUF2156 domain acquired from Pfam [25] (in later revisions renamed LPG synthase C). Starting with this MSA, it was processed to include only amino acid characters and gaps, excluding sequences with ambiguous or non-standard characters. Additionally, sequences with contiguous strings of gaps greater than 20% of the full sequence’s length were removed. This left 11,507 sequences with a length of 298. After the data cleaning process described in Figure 3, a total of 7,890 sequences were used to train our final model.

Model training

The model previously described [14] was utilized, specifically the Python 2.7 version freely available on Github (https://github.com/jertubiana/ProteinMotifRBM). Extensive testing of the combinations of the number of hidden units, L2/L1 regularization, and learning rate were performed. Models were trained in triplicate with different random seeds, as the training procedure converges to slightly different weight values depending on their initial random configuration. Training was considered successful when the model consistently produced similar sets of weights across different random seeds, and the weights had the previously mentioned sparsity and magnitude indicating compositional representation. The final model used in this work was trained for 2000 epochs with a learning rate of 0.1, a learning rate decay of 0.33, L2/L1 parameter set to 1.0, and 300 hidden units.

Model Weight selection

To find weights which were relevant to function we computed the L1 norm of each weight individually, then ranked them. The weights with the largest magnitude were looked at first, and these were typically the sparsest of all weights. Weights were chosen which involved sequence coordinates implicated in our function of interest. Specifically, locations identified through Autodock [26] where the lipid was likely to interact, and a small radius around this region to select a small set of coordinates.

Software Accessibility

DUF2156 domain sequences were retrieved from Pfam33 [25]. Raw sequence datasets are available as Supplementary Material.

eLife Assessment

This study reports important findings on identifying sequence motifs that predict substrate specificity in a class of lipid synthesis enzymes. It sheds light on a mechanism used by bacteria to modify the lipids in their membrane to develop antibiotic resistance. The evidence is convincing, with a careful application of machine learning methods, validated by mass spectrometry-based lipid anlaysis experiments. This interdisciplinary study will be of interest to computational biologists and to the community working on lipids and on enzymes involved in lipid synthesis or modification.

Reviewer #1 (Public Review):

The basic approach is that the authors first train an RBM on all MprF sequences, and then use this analysis to identify a subset of the family that catalyzes the addition of amino acids to PG. Then a second RBM is trained on this subset.

In the initial RBM training a particular hidden unit is identified that has a sparse and bimodal activation in response to the input sequences. The contribution of individual resides is shown in Figure 3c, which highlights one of the strengths of this RBM implementation - it is interpretable in a physically meaningful way. However, there are several decisions here, the justification of which is not entirely clear.

i) Some of the residues in Fig 3c are stated as "relevant" for aminoacylated PG production. But is this the only such hidden unit? Or are there others that are sparse, bimodal, and involve "relevant" AA?

ii) In order to filter the sequences for the second stage, only those that produce an activation over +2.0 in this particular hidden unit were taken. How was this choice made?

iii) How many sequences are in the set before and after this filtering? On the basis of the strength of the results that follow I expect that there are good reasons for these choices, but they should be more carefully discussed.

iv) Do the authors think that this gets all of the aminoacylated PG enzymes? Or are some missed?

The authors show that they can classify members of the family by training a second RBM on the filtered sequences. They do this by identifying two hidden unit activations in particular (Figure 5b) which seem to be useful for determining lipid substrate specificity, and they test several variants that obtain different responses of these two hidden units by experimentally determining what lipids they produce (Table 2). However, some similar criticisms from the last point occur here as well, namely the selection of which weights should be used to classify the enzymes' function. Again the approach is to identify hidden unit activations that are sparse (with respect to the input sequence), have a high overall magnitude, and "involve residues which could be plausibly linked to the lipid binding specificity."

i) Two hidden units are identified as useful for classification, but how many candidates are there that pass the first two criteria? Indeed, how many hidden units are there?

ii) The criterion "involve residues which could be plausibly linked to the lipid binding specificity" is again vague. Do all of the other candidate hidden units *not* involve significant contributions from substrate-binding residues? Maybe one of the other units does a better job of discriminating substrate specificity. (As indicated in Figure 8, there are examples of enzymes that confound the proposed classification.) Why combine the activations of two units for the classification, instead of 1 or 3 or...?

Reviewer #2 (Public Review):

In "Lipid discovery enabled by sequence statistics and machine learning" Christensen et al. address an important question: how can bacteria modify lipid charges to produce cationic lipids, prone to confer resistance to cationic antibiotics? One of the enzymes involved in this process is MprF, which can, through the transfer of amino acids, in particular, lysine, from charged tRNA modify the charge of anionic membrane phospholipid from negative to positive. Recent works have shown that MprF can also modify another substrate, glycolipid glucosyl-diacylglycerol, which is neutral. These findings immediately raise two questions: what are the determinants in the MrpF sequence controlling the lipid substrates it can modify? Are there other substrates for MrpF, so far unknown?

Christensen et al. address both of these questions in an elegant way, combining sequence analysis with machine-learning methods and experimental characterisation of the enzymatic products through mass spectrometry. Using restricted Boltzmann machines (RBM), an unsupervised architecture extracting statistical features from the sequence data, they identify putative amino-acid motifs along the MprF sequences possibly related to the substrate identity, select some bacterial species whose wild-type sequence contains those motifs, and validate the biological role of the motifs by identifying the produced lipids. Remarkably, with this approach, the authors find a novel cationic lipid with two glucosyl groups.

Besides these new results on MrpF and its operation, the present work is appealing, as it shows that the functional characterisation of a very small number of proteins (here, three!) combined with the guided classification of homologous sequence data with appropriate machine-learning methods can lead to the discovery of new functionalities.

Reviewer #3 (Public Review):

Summary:

After the previous identification that the Streptococcus agalactiae MprF enzyme can synthesize also lysyl-glucosyl-diacylglycerol (Lys-Glc-DAG), besides the already known lysyl-phosphatidylglycerol (Lys-PG), the authors aim for the current manuscript was to investigate the molecular determinants of MprF lipid substrate specificity in a variety of bacterial species.

Strengths:

- In general, the manuscript is well constructed and easy to follow, especially taking into account the multidisciplinary aspect of it (computational machine learning combined with lipid biology).

-The added value of the Restricted Boltzmann machines (RBM) approach, in comparison to standard computational pairwise sequence statistics, becomes evident. This is exemplified by a successful, although not perfect, classification and categorization of MprF activity.

- The MS analysis (monoisotopic mass, plus fragmentation pattern), convincingly shows the identification of a novel lipid species Lys-Glc2-DAG.

Weaknesses:

-In many of the analyzed strains, the presence of the lipid species Lys-PG, Lys-Glc-DAG, and Lys-Glc2-DAG is correlated to the presence of the MprF enzyme(s), but one should keep in mind that a multitude of other membrane proteins are present that in theory could be involved in the synthesis as well. Therefore, there is no direct evidence that the MprF enzymes are linked to the synthesis of these lipid species. Although, it is unlikely that other enzymes are involved, this weakens the connection between the observed lipids and the type of MprF.

-Related to this, in a few cases MprF activity is tested, but the manuscript does not contain any information on protein expression levels. Heterologous expression of membrane proteins is in general challenging and due to various reasons, proteins end up not being expressed at all. As an example, the absence of activity for the E. faecalis MprF1 and E. faecium MprF2 could very well be explained by the entire absence of the protein.

Overall, the authors largely achieved their goals, as the applied RBM approach led to specific sequence determinants in MprF enzymes that could categorize the specificity of these enzymes. The experimental data could largely confirm this categorization, although a stronger connection between synthesized lipids and enzyme activity would have further strengthened the observations.

The work now focuses only on MprF enzymes, but could in theory be expanded to other categories of lipid-synthesizing enzymes. In other words, the RBM approach could have an impact on the lipid synthesis field, if it would be a tool that is easily applicable. Moreover, the lipids synthesized by MprF (Lys-PG, but also other cationic lipids) play an important role in bacterial resistance against certain antibiotics.

Author response:

The following is the authors’ response to the original reviews.

Reviewer 1 (Public Review):

The contribution of individual resides is shown in Figure 3c, which highlights one of the strengths of this RBM implementation - it is interpretable in a physically meaningful way. However, there are several decisions here, the justification of which is not entirely clear.

i) Some of the residues in Fig 3c are stated as "relevant" for aminoacylated PG production. But is this the only such hidden unit? Or are there others that are sparse, bimodal, and involve "relevant" AA?

Thanks for bringing this important question to our attention. In fact, this was the only hidden unit involving the combination of positions 152 and 212. Although we don't have knowledge of all relevant amino acids for this catalytic process, the residues we uncover were however shown through experimental analysis to be critical for the catalytic function of two MprF variants, and thus since our protein of interest involved this function, any domain which did not contain these residues were excluded. We can't rule out that the domains we excluded from further analysis could be performing similar catalytic functions, but we found it unlikely considering the amino acids found in the negative portion of the weight were chemically unlikely to form a complex with the amino acid lysine. We have clarified in the text, that this selection is probably a subset of all important amino acids, however, this selection provided predictive power.

ii) In order to filter the sequences for the second stage, only those that produce an activation over +2.0 in this particular hidden unit were taken. How was this choice made?

The +2.0 was chosen as it ensured that the bimodal distribution was split into two distinct distributions.

iii) How many sequences are in the set before and after this filtering? On the basis of the strength of the results that follow I expect that there are good reasons for these choices, but they should be more carefully discussed.

We started with 11,507 sequences and after filtering we had 7,890 to train our model with. We think this number still maintains robust statistics. This is noted in the Dataset acquisition and pre-processing section of the Methods section.

iv) Do the authors think that this gets all of the aminoacylated PG enzymes? Or are some missed?

This is an interesting question that prompted us to do further analysis. We have added a new supplemental figure providing more details to this question. Based on the Uniprot derived annotations and the Pfam domain-based analysis of these sequences, the large majority of sequences that were excluded were proteins which included the LPG_synthase_C domain but not the transmembrane flippase domain required by the MprF class of enzymes, and were instead accompanied by different domains which seem less relevant to our enzyme of interest. It is true though, and related to question (i), that variants which might retain the functionality despite losing experimentally determined key catalytic residues could have been excluded by this method, but such sequences could still be reasonably excluded due to their dissimilarity with MprF from Streptococcus agalactiae.

However, some similar criticisms from the last point occur here as well, namely the selection of which weights should be used to classify the enzymes' function. Again the approach is to identify hidden unit activations that are sparse (with respect to the input sequence), have a high overall magnitude, and "involve residues which could be plausibly linked to the lipid binding specificity."

(i) Two hidden units are identified as useful for classification, but how many candidates are there that pass the first two criteria? Indeed, how many hidden units are there?

We note in the Model training section of the methods that our final model used had 300 hidden units in total. As to the first part of your question, rather than systematically test the predictive power of all other hidden units to this task, we decided to use the weights that we did because of their connection to a proposed lipid binding pocket found through Autodocking experiments. While another weight might provide predictive power, it might lack this critical secondary information. Moreover, the direction of our research necessitated finding weights which first satisfied our lipid-binding pocket plausibility before using these weights to propose MprF variants to test for our novel functionality. Given the limited information we had early in the research process, to go in reverse would have provided too many options for experimental testing with reduced mechanistic justification. We included a brief explanation of our rationale in section " Restricted Boltzmann Machines can provide sensitive, rational guidance for sequence classification “ in the updated manuscript.

ii) The criterion "involve residues which could be plausibly linked to the lipid binding specificity" is again vague. Do all of the other candidate hidden units *not* involve significant contributions from substrate-binding residues? Maybe one of the other units does a better job of discriminating substrate specificity. (As indicated in Figure 8, there are examples of enzymes that confound the proposed classification.) Why combine the activations of two units for the classification, instead of 1 or 3 or...?

In fact, it is true that the other hidden units do not involve significant contribution to substrate-binding residues, and we will clarify this. The weights found through this RBM methodology are biased to be probabilistically independent, meaning that the residues and amino acids implicated by each weight are not shared among the other weights through the design of the model. We will update the Model Weight selection section to clarify that the weights we chose had more significantly weighted residues overlapping with the residues near the lipid-binding region than the other weights we checked. We combined these two because they were the only ones which had both overlap with these residues and predictive power of lipid activity with the few sequences we had detailed knowledge of at the time of decision (Figure 5b).

The Model Weight section reads as follows:

“Weights were chosen which involved sequence coordinates implicated in our function of interest. Specifically, locations identified through Autodock (Hebecker et al., 2015) where the lipid was likely to interact, and a small radius around this region to select a small set of coordinates. We chose the only weights which had both overlap with multiple residues in this chosen radius and predictive power (separation) for the three examples we had to start with.”

Author Recommendations:

The manuscript will likely be read by many membrane biologists/biochemists, and they might like to better understand how the RBM might be useful in their own approach. Here are some suggestions along these lines. The overall goal is to explain the RBM in *plain English* - the mathematical description in Eqs 2-4 is not easily interpretable.

(1a) Explain that the RBM is a two-layer structure, in which one layer is the "visible" elements of the input sequence, and the other is called "hidden units." Connections are only made between visible and hidden units, but all such connections are made.

(1b) The strengths of these connections are called "weights", and are determined in a statistical way based on a large set of input sequences. Once parametrized, the RBM is capable of capturing correlations among many positions in an input sequence - a significant advantage over the DCA approach.

We agree with this assessment, and have updated the section of the text where we introduce the RBM with a non-technical explanation of what this method is doing. It reads as:

“The design of this RBM can be seen in Figure 4, where the model architecture is represented by purple dots and green triangles. The dots are the “visible” layer, which take in input sequences and encode them into the “hidden” layer, where each triangle represents a separate hidden unit. The lines connecting the visible and hidden layers show that each hidden unit can see all the visible units (the statistics are global), but they cannot see any of the other hidden units, meaning the hidden units are mutually independent. This global model with mutually independent hidden units (see also the marginal distribution form shown in Equation 3) has the following useful properties: higherorder couplings between... “

(1c) Although strictly true that the DCA model is a Boltzmann machine, it's not a typical Boltzmann machine, because all of the units are visible. Typically a Boltzmann machine would also include hidden units, in order to increase its capacity/power.

We have clarified the relationship between DCA and Boltzmann machines, and this section now reads as:

This class of models is closely related to another model termed the Boltzmann machine. The Boltzmann machine formulation is closely related to the Potts model from physics, which was successfully applied in biology to elucidate important residues in protein structure and function (Morcos et al., 2011), and another example being the careful tuning of enzyme specificity in bacterial two-component regulatory systems (Cheng et al., 2014; Jiang et al., 2021). The Boltzmann machine-like formulation from Morcos et al. (2011), termed Direct Coupling Analysis (DCA), stores patterns...

(1d) Throughout, the authors refer to the activation of the hidden units as weights, but this is not a typical usage of this terminology. Connections between units are weights and have two subscripts. Given an input sequence, the sum over these weights for a given hidden unit is its activation (Eq. 1). I suggest aligning the description with the typical usage in order to make the presentation easier to follow. Hereafter I will refer to these hidden unit activations as simply activations.

We agree with you, the hidden units are a collection of edge weights. We have modified the terminology in the text and in our figures to consistently refer to the collections of weights as hidden units and refer to the hidden unit outputs given a sequence input as activations.

(1e) How many hidden units are there?

The final model was trained with 300 hidden units.

(2) It is redundant to say that lipids are both amphiphiles and hydrophobic...amphiphile already means hydrophobic plus hydrophilic.

This is true, we have edited the manuscript to reflect this.

(3) What does this mean, and what's the point of this remark? "They [lipids] are relatively smaller than other complex biomolecules, such as proteins, thereby allowing a larger portion of their surface to interact with other macromolecules."

We have removed this sentence.

Reviewer 2 (Author Recommendations):

While the idea of filtering out a part of the sequence data obtained with BLAST makes sense per se, it would be nice if the authors could comment on the nature of the sequences corresponding to the left peak in Figure 3b. It is hypothesised in conclusion that these sequences could lack any catalytic function. Could the authors experimentally check that this is the case or provide further evidence for this hypothesis?

Yes, in this revision we provide further evidence as a new supplementary figure S2. At the time we performed domain analysis of the sequences we excluded; most of these sequences lacked the flippase domain associated with MprF function, and instead were combined with different domains. On this basis we excluded them due to their lack of relevance to the MprF from Streptococcus agalactiae we were interested in. Although there is possibility that some relevant sequences might be excluded, our assessment is that we gained specificity by reducing the set of sequences.

A key step in the RBM-based approach is the identification of "meaningful" hidden units, i.e. whose values are related to biological function. In Methods, the authors explain how they selected these units based on the L1 norms of the weights and the region of interaction with the lipid. While these criteria are reasonable, I wonder whether they are too stringent. In particular, one could think that regions in the proteins not in direct contact with the lipid could also be important for binding. It is known for instance that the length of loops can affect flexibility and help regulate activity in some catalytic enzymes. So my question is: if one relaxes the criterion about the coordinates of large weight values, what happens? Are other potentially interesting hidden units identified?

We completely agree that other regions of the protein are likely involved in determining enzyme specificity, and that focusing on solely regions which interact with the lipid is perhaps missing important contributions to the catalytic function; we hypothesize that the flippase domain itself and its interaction with the catalytic domain are involved, especially considering the concerted mechanism by which they must operate. We are currently investigating these theories and will be the subject of future work. As an initial step, we present this current work with restricted information that led to concrete predictions. We focused on the lipid binding pocket because it was one of just a few bits of information we had from the start, but as the reviewer suggests, we plan to follow up our research to try to identify other relevant hidden units and domains.

From a purely machine-learning point of view, it would be good to see more about cross-validation of the model. More precisely, could the authors show the log-likelihood of test set data compared to the one of training sequence data?

We agree this is an important piece of information. We will update our methods section with this information. We performed a parameter sweep to search for the parameter’s we used in our final model, and in that testing with a random 80/20% training/test split we had a training log probability loss of -0.91, and a test loss of -0.98. However, for our final model we used all available data and did not perform a split; the final result did not change dramatically by including the additional data, and the weight structure and composition was consistent with the results presented in the paper.

Reviewer 3 (Public Review):

In many of the analyzed strains, the presence of the lipid species Lys-PG, Lys-Glc-DAG, and Lys-Glc2-DAG is correlated to the presence of the MprF enzyme(s), but one should keep in mind that a multitude of other membrane proteins are present that in theory could be involved in the synthesis as well. Therefore, there is no direct evidence that the MprF enzymes are linked to the synthesis of these lipid species. Although, it is unlikely that other enzymes are involved, this weakens the connection between the observed lipids and the type of MprF.

While there are a number of proteins found on the membrane that could play a role, we have specifically used a background strain that has a transposon in mprF that makes the bacteria incapable of synthesizing Lys-lipids (Figure 7B) unless complemented back with a functional MprF (Figure 7D-E). This led us to conclude that MprF is responsible for Lys-lipid synthesis.

Related to this, in a few cases MprF activity is tested, but the manuscript does not contain any information on protein expression levels. Heterologous expression of membrane proteins is in general challenging and due to various reasons, proteins end up not being expressed at all. As an example, the absence of activity for the E. faecalis MprF1 and E. faecium MprF2 could very well be explained by the entire absence of the protein.

The genes were expressed on the same plasmid to control for expression. While we did not run a western blot to examine expression levels the plasmid backbone was used as a control for protein expression. Previous research supports E. faecalis MprF1 and E. faecium MprF2 not synthesizing Lys-lipids and instead most likely play a different role in the cell membrane.

The title is somewhat misleading. The sequence statistics and machine learning categorized the MprFs, but the identification of a novel lipid species was a coincidence while checking/confirming the categorization.

We believe the title is appropriate given that the identification of Enterococcus dispar was through computational methods that led to the discovery Lys-Glc2-DAG. In other words, the categorization of potential organisms that produce lipids related to MprF has been driven by the proposition from the computational method. We agree, however, that the discovery was unexpected but would not have happened without the suggested organisms coming from the methodology presented here.

Please read the manuscript one more time to correct textual errors.

The example of the role of LPS in delivering siRNA to targeted cancer cells is a bit farfetched as LPS is very different from the lipids that are being discussed here. I would rather focus on the role of Lysyl-lipids in antibiotic resistance in the introduction.

We included LPS here to explain that natural lipids/components of the bacterial cell membrane could be used for drug delivery systems. While it is true LPS is quite different from Lys-lipid compounds, our goal was to create an emphasis on how the bacterial domain is a rich untapped source of lipids that could be used in biotechnology. In this way we wanted our statement to be more broadly about bacterial lipids and the importance of their continued study for diverse applications like pharmaceuticals.

The MS identification of Lys-Glc2-DAG is convincing, especially in combination with the fragmentation data, but the ion counts suggest low abundance. The observation would be strengthened if the identification of Lysyl-Glc2-DAG with different acyl-chain configurations has been observed. This should be then mentioned or visualized in the manuscript.

We agree and have added an updated Figure 8A to demonstrate the presence of different acyl-chain configurations in Enterococcus dispar.

Further analysis of the Enterococcus strains shows the presence of the three lipids Lys-PG, Lys-Glc-DAG, and LysGlc2-DAG, although the Lys-Glc-DAG is only detected in trace amounts. This raises questions on the specificity of the MprF for the substrate Glc-DAG. If the ratio of Glc2-DAG compared to Glc-DAG abundance is similar to the ratio of Lys-Glc2-DAG vs. Lys-Glc-DAG abundance, this would strengthen the observation that the enzyme has equal affinity. However, if there is a rather large amount of Glc-DAG but a small amount of Lys-Glc-DAG, the production of Lys-Glc-DAG might be a side-reaction.

The reviewer brings a relevant point of discussion, however, a clear resolution might be part of future work as we do not use spike in controls when completing lipid extractions. Because of this, it it is not possible for us to compare lipid levels across different samples. We now include a note clarifying this in the discussion section.

The plotting of the MprF sequence variants using the chosen RBM weights reveals a rather complex distribution over the quadrants (Figure 8). It is rather unclear in Figure 8 why only 1 sequence is plotted for Enterococcus faecalis and faecium, while 2 different MprFs are present (and tested) for these two organisms. This should be clarified.

We agree this can be a source of confusion. We have further clarified this in the text that only the functional alleles were plotted in Figure 8 and that all Enterococcal alleles are plotted in Figure S3 regardless of function.

Version 2

Abstract

Abstract

Bacterial membranes are complex and dynamic, arising from an array of evolutionary pressures. One enzyme that alters membrane compositions through covalent lipid modification is MprF. We recently identified that Streptococcus agalactiae MprF synthesizes lysyl-phosphatidylglycerol (Lys-PG) from anionic PG, and a novel cationic lipid, lysyl-glucosyl-diacylglycerol (Lys-Glc-DAG), from neutral glycolipid Glc-DAG. This unexpected result prompted us to investigate whether Lys-Glc-DAG occurs in other MprF-containing bacteria, and whether other novel MprF products exist. Here, we studied protein sequence features determining MprF substrate specificity. First, pairwise analyses identified several streptococ-cal MprFs synthesizing Lys-Glc-DAG. Second, a restricted Boltzmann machine-guided approach led us to discover an entirely new substrate for MprF in Enterococcus, diglucosyl-diacylglycerol (Glc2-DAG), and an expanded set of organisms that modify glycolipid substrates using MprF. Overall, we combined the wealth of available sequence data with machine learning to model evolutionary constraints on MprF sequences across the bacterial domain, thereby identifying a novel cationic lipid.

Introduction

Lipids are ubiquitous and diverse group of amphiphathic compounds that are critical for fundamental biological processes, including the formation of cell membranes, protection, and cargo delivery; energy storage; and cell signaling pathways [1]. Chemical modifications to lipids, including changes in fatty acid tails (saturated and unsaturated) [2], modifications of head groups [3], and alterations to lipid charge [4], impact these interactions, as well as the physicochemical properties of membranes. Commonly, bacteria synthesize lipids that are negatively charged or neutral [5]. Some bacteria modify the negatively charged lipids with amino acids to make them positively charged to confer resistance to cationic antimicrobial peptides and cationic antibiotics [6, 7]. The bacterial domain is a rich source of natural lipids that are likely co-evolved for specific interactions with host tissues and other cells, yet the full extent of this chemical diversity remains unexplored. Finding bacteria that possess unique lipids can be useful for biotechnology. For example, natural lipids found in bacteria can be used for drug delivery. A study by [8] used low cytotoxic outer membrane vesicles (OMV) that contained a modified lipopolysaccharide (LPS) to deliver siRNA to targeted cancer cells. This delivery mechanism had success and demonstrates the possibility of using naturally occurring bacterial lipids for future drug delivery processes to eukaryotic cells.

One enzyme responsible for modifying lipids in bacteria is the multiple peptide resistance factor (MprF). MprF modifies lipids through the transfer of amino acids from charged tRNAs to the head group of the anionic membrane phospholipid, phosphatidylglycerol (PG) [4]. The MprF protein is comprised of two domains: an N-terminal flippase domain responsible for moving aminoacylated lipids from the inner membrane leaflet to the outer leaflet and a C-terminal synthase domain responsible for the aminoacyl transfer from tRNA to lipids [9, 10]. Recently, in Streptococcus agalactiae (Group B Streptococcus, GBS) it was found that MprF can modify two different substrates with lysine (Lys) – the glycolipid glucosyl-diacylglycerol (Glc-DAG) and the phospholipid, PG [11], generating Lys-Glc-DAG, as well as Lys-PG [11, 12] (Figure 1). An S. agalactiae mprF deletion mutant no longer synthesizes Lys-Glc-DAG or Lys-PG and expression of S. agalactiae mprF in the heterologous host Streptococcus mitis conferred Lys-Glc-DAG and Lys-PG synthesis to S. mitis [11]. This was the first time MprF was demonstrated to add lysine onto a neutral glycolipid (Glc-DAG).

Chemical structures of lysine lipids. Lys-PG, lysyl-phosphatidylglycerol; Lys-Glc-DAG, lysylglucosyl-diacylglycerol; Lys-Glc2-DAG, lysyl-diglucosyl-diacylglycerol.

The synthesis of Lys-Glc-DAG by S. agalactiae MprF illustrated a unique lipid substrate and product by the enzyme that had not been characterized before, highlighting the possibility for further exploration into unknown lipids and lipid substrates by MprF of different bacterial species. We sought to investigate the molecular determinants of enzyme specificity and identify other bacterial species that may have this novel specificity. This investigation led us from the standard methodology of BLASTp queries [13], which are based on single residue amino acid substitution frequencies, to a more modern statistical method called a Restricted Boltzmann Machine (RBM) [14], which captures global statistical patterns within natural sequence data. In this work, we develop a general strategy for combining this high-powered statistical model with lipidomic analysis to discover MprF variants with unique enzymatic activity. This process led us to the discovery of an uncharacterized lipid substrate for MprF, diglucosyl-diacylglycerol (Glc2-DAG), which is present in several strains identified by the parameters of the statistical model. These models also provide interpretable and actionable information for use in further enzyme characterization, and we show that this application is a useful companion tool for experimental work.

Results

Simple pairwise sequence statistics identify streptococcal MprF enzymes that synthesize Lys-Glc-DAG

Utilizing a method based on the amino acid sequence of S. agalactiae MprF, we identified four streptococcal MprFs with high sequence identity to S. agalactiae COH1 MprF (WP 000733236.1). A BLASTp analysis found that Streptococcus sobrinus ATCC 27352 MprF (WP 019790557.1) had 61.4% amino acid identity to the S. agalactiae MprF, Streptococcus salivarius ATCC 7073 MprF (WP 002888893.1) had 43.09% amino acid identity, Streptococcus ferus MprF (WP 018030543.1) had 61.88% amino acid identity, and Streptococcus downei (WP 002997695.1) had 61.03% amino acid identity (Table 1). Plasmids were designed to express the S. sobrinus mprF (pSobrinus), S. salivarius mprF (pSalivarius), S. ferus mprF (pFerus), and S. downei mprF (pDownei) genes. The plasmids were transformed into the expression host S. mitis NCTC12261 (SM61). The empty vector (pABG5ΔphoZ, referred to as pABG5 here) was also transformed (Figure 2a). S. mitis does not natively encode mprF but synthesizes lipids (Glc-DAG; PG) that are substrates for S. agalactiae MprF [15, 16, 11], making it an appropriate heterologous host for expressing streptococcal MprFs. Previously, a plasmid expressing S. agalactiae mprF (pGBSMprF) [11] was generated and transformed into S. mitis. Lys-Glc-DAG and Lys-PG presence in SM61 was only possible with the expression of pGBSMprF (Figure2b) [11].

Synthesis of lysine lipids (Lys-PG and Lys-Glc-DAG) in S. mitis expressing mprFs from S. agalactiae, S. salivarus, and S. ferus. (a) S. mitis NCTC12261 with empty vector control (pABG5) lacks lysine lipids; (b) S. agalactiae mprF (pGBSMprF) produces both Lys-PG and Lys-Glc-DAG; (c) S. salivarius mprF produces only Lys-PG; (d) S. ferus mprF produces only Lys-Glc-DAG. Left panels: total ion chromatograms (TIC); middle panels: mass spectra of retention time 19.5–21.5 min showing Lys-PG and PC; right panels: mass spectra of retention time 26–30 min showing Lys-Glc-DAG. Note: “*” is an extraction artifact due to chloroform used. DAG, diacylglycerol; MHDAG, monohexosyldiacylglycerol; DHDAG, dihexosyldiacylglycerol; PG, phosphatidylglycerol; Lys-PG, lysyl-phosphatidylglycerol; Lys-Glc-DAG, lysylglucosyl-diacylglycerol; PC, phosphatidylcholine.

S. mitis lipids of strains expressing the plasmids pSobrinus (Table 1), pSalivarius (Figure 2c), pFerus (Figure 2d) and pDownei (Table 1) were analyzed. Lipidomic analysis found that pSobrinus, pDownei, and pFerus conferred synthesis of Lys-Glc-DAG, but not Lys-PG in S. mitis. Notably, pFerus confers synthesis of Lys-Glc-DAG at a similar level to S. agalactiae MprF (Figures 2b, d). In contrast, pSalivarius conferred synthesis of Lys-PG only and not Lys-Glc-DAG. Importantly, S. salivarius MprF also synthesized Lys-PG at a level most similar to S. agalactiae MprF (Figures 2b, c). Lysine-modified lipids synthesized by S. mitis strains expressing streptococcal MprFs used in this study are summarized in Table 1. In general, streptococcal MprF enzymes with higher amino acid percent (> 60%) to S. agalactiae MprF conferred Lys-Glc-DAG synthesis to S. mitis, but only S. agalactiae MprF conferred both Lys-Glc-DAG and Lys-PG synthesis in S. mitis.

Restricted Boltzmann Machines can provide sensitive, rational guidance for sequence classification

We set out to understand which specific MprF amino acid residues are involved in enzyme specificity for either the PG or Glc-DAG substrate. To this end, we employed a Restricted Boltzmann Machine (RBM), a probabilistic graphical model which aims to model the probability of data in a data set through statistical connections between the features of the data and a set of hidden units. [17]. This class of models is closely related to another model termed the Boltzmann Machine. The Boltzmann machine formulation is closely related to the Potts model from physics, which was successfully applied in biology to elucidate important residues in protein structure and function [18], one example being the careful tuning of enzyme specificity in bacterial two-component regulatory systems [19, 20]. The Boltzmann Machine formulation from [18], termed Direct-Coupling Analysis (DCA), stores patterns in the form of a pairwise coupling matrix and local field matrix (shown in Equation 2), which has been useful in understanding how likely pairs of residues are to interact in physical space. We hypothesize that S. agalactiae MprF and other MprF enzymes that utilize glycolipid substrates have unique patterns of interacting residues relative to those enzymes that use only PG as a substrate.

While modeling how pairs of residues covary is powerful, many important features of enzyme function like allostery or enzyme specificity (which is the purpose of this analysis) may involve more than two residues [21], and in these situations it can be difficult to connect sets of residues using solely pairwise parameters [22]. Therefore, a more global approach is required to address our hypothesis. To this end, a RBM has recently been developed for use in the learning of protein sequence data, with special attention given to producing a model which produces sparse, interpretable, and biologically meaningful representations of these higher-order statistical couplings within the hidden units [14]. The design of this RBM can be seen in Figure 3, where the model architecture is represented by purple dots and green triangles. The dots are the “visible” layer, which take in input sequences and encode them into the “hidden” layer, where each triangle represents a separate hidden unit. The lines connecting the visible and hidden layers show that each hidden unit can see all the visible units (the statistics are global), but they cannot see any of the other hidden units, meaning the hidden units are mutually independent. This global model with mutually independent hidden units (see the marginal distribution form shown in Equation 3) allows the following useful properties: higher-order couplings between residues (sets of coupled residues instead of pairs); bimodal hidden unit outputs given the training data set (Equation 4); sparse, interpretable configurations of the hidden units (wμ) (Equation 5) and a compositional representation of the hidden unit weights [23], where input sequences are modeled largely through combinations of hidden units instead of single, highly activating units.

Schematic of the RBM methodology. An aligned set of protein sequences is first used to learn a hidden unit representation that best describes the statistics of the sequence dataset given restrictions on the hidden unit representation. Then, the individual hidden units can be studied to find particular units which allow useful enzyme classification, and additionally, these weights can be meaningfully interpreted as statistically co-varying sequence configurations. Additionally, the classification can be used to create filtered datasets to train more models.

The compositional representation is a critical feature of this method, allowing us to find independently coupled sets of residues which in combination describe these protein sequences, and it is achieved through the particular choices of training parameters. More details on the training parameter choices are found in Methods and Materials. In particular, we focus on the scoring of sequences using individual hidden units wμ: where v is an input sequence vectorized through one-hot encoding, to produce a hidden unit activation (a single number) which depends on the input sequence residues at position i and the value of the hidden unit weight matrix for that residue at that position.

We used this methodology to study the sequences in the family of MprF, which is a large protein composed of independent flippase and synthase domains [9, 24]. In our analysis, we restrict ourselves to the cytosolic region (see yellow box in Figure 4a), also known as the aminoacyl-PG synthase domain, which constituted the Pfam family DUF2156 (now renamed LPG synthase C) [25]. One hidden unit in particular is highlighted, where we find bimodal activation (Figure 4b) of the hidden unit (Figure 4c) when Equation 1 is evaluated with the training sequences. Interpreting the sequence histograms, a positive value activation in Figure 4b corresponds to a sequence containing combinations of the residues listed in the positive portion of the weight diagram in Figure 4c, whereas a negative activation corresponds to residues in the negative portion. For display purposes, the residue weights shown in all of our hidden units correspond to the positions where the magnitude of the visible positions wiμ are greater than some threshold proportional to the largest magnitude position (position 212); the hidden units has values for all possible amino acids in the entire protein length, but they are considerably smaller than this threshold and generally very low.

Example of hidden unit analysis and usage. (a) The structure of PDB:7DUW, with the red colored region being the transmembrane flippase domain and the yellow boxed region the cytosolic domain which we focus on. (b) The activations produced by inputting a sequence into a hidden unit, producing a single number as output which corresponds to a summation of negatively and positively weighted residues. Performed on entire training set (histogram in blue), highlighting sequences corresponding to predominantly positive weighted residues. (c) Hidden unit from an RBM trained on the Pfam DUF2156 domain. The MSA positions 152 and 212 correspond to residues S684 and R742, respectively. (d) Residues (in yellow) in the MprF cytosolic domain which form the binding pocket for Lys-tRNALys (the ligand analogue L-lysine amide shown in green), from PDB:4V36. LYN, L-lysine amide.

Two positions in particular, 152 and 212, in the positive portion of the hidden unit correspond to residues 684 and 742, which were experimentally demonstrated to be relevant for aminoacylated-PG production by the Bacillus licheniformis and Pseudomonas aeruginosa cytosolic synthase domain variants [26]; in B. licheniformis the residues which form a complex with the lysine from Lys-tRNALys are shown in Figure 4c. We have highlighted these two aforementioned variants as well as the S. agalactiae variant for demonstration of sequence scoring through Equation 1 in Figure 4a. We used this hidden unit to create a filtered data set; only sequences that contained this necessary motif were of interest to us, and we subsequently trained a new model using only sequences with a catalytic hidden unit activation greater than 2 (because it cleanly split the bimodal distribution). We outline this general workflow in Figure 3. We also note that the full length sequences of the sequences included in our new dataset are enriched with sequences which contain the bacterial flippase domain (shown in red in Figure 4a), and the sequences we removed rarely contain the flippase domain (see Figure S2 for domain analysis).

RBM hidden units identify plausible sequence residues for functional characterization

After filtering the data set and training another model, we set out to find hidden units which could describe MprF specificity. To this end, we identify hidden units which are sparse, have high overall magnitude, and involve residues which could be plausibly linked to the lipid binding specificity (see Methods for more details). Through this search method we identified the two hidden units shown in Figure 5a. They both have high sparsity and high magnitude relative to the majority of other hidden units in the model, and importantly they correspond to residues localized to regions of the catalytic domain that had been previously identified through Autodock experiments using a PG molecule with side chains C5:0/C8:0 [26] (highlighted in Figure 5a). The scoring of the training set of sequences with these hidden units is shown in Figure 5b, and we see a clear separation between the cytosolic domain sequences which produces Lys-Glc-DAG (from S. agalactiae) and two domain variants which produce only aminoacylated-PG (from P. aeruginosa and B. licheniformis). Importantly, these output activations depend on a greatly reduced subset of the full sequence and do not correspond to clustering on total sequence identity (Figure S1).

Proposed set of hidden units for classifying lipid specificity. (a) two hidden units found in an RBM trained on the filtered Pfam dataset. The hidden unit residues are highlighted in the PDB:4V34 structure, with arrows pointing to their corresponding residue sets. (b) the activations of the hidden units when scoring the sequences used in the training set and sequences from NCBI not used during training (N=23,138). S. agalactiae produces Lys-Glc-DAG and Lys-PG, while B. licheniformis and P. aeruginosa produces only aminoacylated-PG. Q1-Q4 are quadrant labels which we refer to throughout the paper.

Using these two hidden units as sequence classifiers with a potential link to the substrate specificity of cytosolic domain variants, we used Figure 5b to propose sequences in quadrant three (Q3), the quadrant occupied by the S. agalactiae MprF variant with this novel glycolipid-modifying function. We selected a sequence which was distant in terms of sequence identity from the Streptococcus genus (Table S1) and which has been implicated in human pathologies [27], Enterococcus dispar MprF. We also emphasize that the two hidden units we selected are from a set of 300 total hidden units, and that while other hidden units may differentially classify our three sequences, without the additional link to lipid binding provided by the Autodocking experiment there would be only weak justification for using their activations to make predictions.

Enterococcal MprF enzymes synthesize Lys-Glc-DAG, Lys-PG, and a novel cationic glycolipid Lys-Glc2-DAG

Enterococcus dispar ATCC 51266 was identified as a possible candidate for novel lipid synthesis through the analyses described above. Remarkably, our lipidomic analysis found that E. dispar synthesizes a novel cationic lipid, lysyl-diglucosyl(Glc2)-diacylglycerol (Lys-Glc2-DAG) (Figure 1, Figure 6a,b), as well as Lys-Glc-DAG and Lys-PG (Figure 6c, Figure S3). The structural identification of Lys-Glc-2-DAG was supported by exact mass measurement ([M+H]+ observed at m/z 1047.730; calculated m/z 1047.731) and tandem MS (Figure 6b). Lys-Glc-2-DAG is unusually polar and charged for a lipid; chromatographically, Lys-Glc2-DAG is highly retentive on a silica-based normal phase column and elutes off the column at the very end of the gradient (Figure S3). The use of an amino-based HPLC column led to much improved elution profiles for Lys-Glc2-DAG and other glyco- and lysine lipids (Figure 6).

Identification of Lys-Glc2-DAG in E. dispar. (a) Positive ion mass spectrum of Lys-Glc2-DAG species in E. dispar. Major Lys-Glc2-DAG species with carbon atoms (before colon) and double bonds (after colon) (b) MS/MS product ions and fragmentation scheme of the most abundant Lys-Glc2-DAG ion at m/z 1047.73. (c) Extracted ion chromatograms of LC/MS of lysine lipids (Lys-PG, Lys-Glc-DAG, Lys-Glc2-DAG) and Glc2-DAG separated on an amino HPLC column. Glc2-DAG, diglucosyl-diacylglycerol; Lys-PG, lysyl-phosphatidylglycerol; Lys-Glc-DAG, lysyl-glucosyl-diacylgylcerol; Lys-Glc2-DAG, lysyl-diglucosyl-diacylglycerol.

The discovery of Lys-Glc2-DAG in Enterococcus dispar led us to analyze other Enterococcus strains: E. faecium 1,231,410, E. faecalis T11 and OG1RF E. gallinarium EG2, E. casseliflavus EC10, and E. raffinosus Er676. All Enterococcus strains examined were found to synthesize Lys-PG, Lys-Glc-DAG and Lys-Glc2-DAG at varying levels (Figure 7a and Table 2). MprF-dependent Lys-PG production by E.faecalis and E. faecium was previously reported [28, 29], but cationic glycolipids production has not been previously reported. E. gallinarium EG2 and E. casseliflavus EC10 encode a single copy of mprF gene, while the other enterococci we tested encode two distinct mprF genes (mprF1 and mprF2) [28, 29]. When these MprF sequence variants are plotted using our chosen RBM hidden units, we see that the two single copy strains plot in quadrant two (Q2), while for the two copy sequences, each MprF variant maps to separate coordinates (Figure S4). Both E. raffinosus Er676 MprF variants plot to quadrant two (with the lowest Weight 2 value being 0.34). The E. faecium and E. faecalis variants have copies plotting directly in quadrant one and variants plotting farther out from the sequence cluster; we note that the two displaced variants are both lacking the characteristic aspartic acid and glutamic acid amino acids at positions 100 and 48 in the hidden units (Figure 5a), which are the dominant contributions for determining quadrant one occupancy.

E. faecalis MprF2 and E. faecium MprF1 confer Lys-Glc2-DAG synthesis. (a) OG1RF-WT; (b) OG1RF_10760::Tn lacks lysine lipids; (c) OG1RF_10760::Tn + pABG5 lacks lysine lipids; (d) OG1RF_10760::Tn + pOGMprF2 restores lysine lipids; (e) OG1RF_10760::Tn + pEFMprF1 restores lysine lipids; (f) OG1RF_10760::Tn + pEFMprF2 lacks lysine lipids. Expression of OGMprF2 and EFMprF1 in OG1RF_10760 Tn mutant restore Lys-Glc2-DAG synthesis. Shown are the extracted ion chromatograms of lysine lipids (Lys-PG, Lys-Glc2-DAG) and Glc2-DAG separated on an amino HPLC column. Note: Lys-Glc-DAG was found in trace amounts or missing from lipid extractions. Glc2-DAG, diglucosyl-diacylglycerol; Lys-PG, lysyl-phosphatidylglycerol; Lys-Glc2-DAG, lysyl-diglucosyl-diacylglycerol.

To clarify which variants were determinants of the novel cationic lipid production, lipidomic analysis on an E. faecalis OG1RF mini-mariner (EfaMarTn) transposon mutant [30] with a transposon insertion within OG1RF_10760 mprF (referred to as mprF 2 in previous studies [28]; OG1RF_10760::Tn) revealed that Lys-PG, Lys-Glc-DAG and the newly identified Lys-Glc2-DAG were absent when the synthase domain of the gene was disrupted (Figure 7b, Table S1). This led us to the conclusion that E. faecalis mprF 2 is necessary for the synthesis of the three Lys-lipids in E. faecalis. When the empty vector (pABG5) was transformed into OG1RF_10760::Tn, no Lys-lipids synthesis was observed as expected (Figure 7c). This phenotype was complemented by expression of E. faecalis mprF 2 from a plasmid (pOGMprF2) (Figure 7d). Additionally, expression of E. faecium mprF1 (EFTG_00601) (pEfMprF1) restored Lys-PG and Lys-Glc2-DAG synthesis to the E. faecalis OG1RF_10760::Tn strain (Figure 7e), while E. faecium mprF2 (EFTG_02430) did not rescue Lys-lipid synthesis (Figure 7f). Taken together, these data indicate that E. faecalis MprF2 (OG1RF_10760) and E. faecium MprF1 (EFTG_00601), like S. agalactiae MprF, act on both phospholipid and glycolipids substrates, additionally synthesizing the novel cationic glycolipid, Lys-Glc2-DAG.

Hidden unit 2 is correlated with GlcN-DAG substrate activity in MprF

When we started our RBM analysis, we had substrate specificity information for only the quadrants one and three, and through identifying the novel substrate in E. dispar we expanded our testing to quadrant two by testing Enterococcus strains. To expand our RBM analysis, we tested specific sequences which were found in quadrant four (Q4). We conducted lipid analysis on Ligilactobacillus salivarius ATCC 11741, Levilactobacillus brevis ATCC 14869, Lacticaseibacillus rhamnosus ATCC 7469, Lactobacillus casei ATCC 393, and Lacticaseibacillus paracasei ATCC 25302. All strains tested synthesize Lys-PG (Table 2). Only Levilactobacillus brevis and Lacticaseibacillus paracasei synthesized Lys-Glc2-DAG, L. paracasei synthesized low levels of Lys-Glc2-DAG. An additional quadrant one (Q1) strain Exiguobacterium acetylicum UTDF19-27C, was analyzed by lipidomics. It was found to synthesize only Lys-PG. This MprF falls near Staphylococcus aureus RN4220 and Bacillus subtilis 168 in our plot, both of which synthesize only Lys-PG [31, 24] (and experimentally confirmed in this study). This further supports that strains located in quadrant one of Figure 8 are capable of Lys-PG synthesis.

The two proposed hidden unit activations with all of the sequences from Table 2 labeled. Protein ID of highlighted sequences are listed in Table S1. (#) indicates the lipid activity was confirmed through heterologous expression.

Summaries of all functional species/variants tested and the Lys-lipids they produced are shown in Table 2, and plotted in the two RBM hidden units in Figure 8. We list for each MprF variant which of the three tested lipids it acts on, and in Figure 8 we indicate whether a variant has only PG as a substrate or if it operates on GlcN-DAG as a substrate irrespective of its PG activity. With this plot, we can see that sequences which plot with positive activations in hidden unit 2 are more likely to have GlcN-DAG as a substrate (8/9), and sequences with negative values predominantly use PG (7/11), with a Fisher’s exact test p-value of 0.028 (Table S2). Hidden unit 1, however, is not correlated with GlcN-DAG specificity, with a Fisher’s exact p-value of 0.67. Therefore we conclude that hidden unit 2, and thus the high magnitude sequence positions identified by it, are correlated with GlcN-DAG specificity.

Discussion

The utilization of the S. agalactiae MprF amino acid sequence as a query for BLASTp was a simple method to identify various streptococcal MprFs and further our understanding of MprF specificity. This allowed us to identify three MprFs that can synthesize the highly cationic Lys-Glc-DAG. Importantly, although highly similar to S. agalactiae MprF, this method did not identify other streptococcal MprFs capable of lysine addition to both glycolipid and phospholipid substrates (generating Lys-Glc-DAG and Lys-PG) nor did it lead to the identification of a novel cationic glycolipid found in Enterococcus. Sequence logos [32] may identify important sequence features, but largely in instances where amino acids are highly conserved. When amino acids must interact to produce a specific function, yet the exact identity of those amino acids is not a requirement, methods which model the strength of amino acid coevolution in sequence sets can find important signals which in the sequence logo might appear to be negligible. For this reason we used the Restricted Boltzmann Machines (RBM) method [14], which can model statistical connectivity of residues at multiple sites in the protein sequence.

The RBM can be used in a number of ways to study sequence data. Firstly, it allowed us to filter the sequence data using hidden units (Figure 4); the two hidden units which formed the basis of our specificity analysis were not found in models trained on the Pfam dataset when it was initially pulled from Pfam, and this was potentially due to the presence of sequences without relevant catalytic function adding noise to the dataset. The most prominent feature for our purposes is the ability to use the hidden units learned through RBM training to classify sequences. The bimodal nature of this particular model provides a relatively simple interpretation which is straightforward to use for prediction. We show in Figure 5 that the combination of clustering and interpretable hidden unit structure allows us to identify a small subset of residues within a structure, grounding our statistical clustering in MprF’s structural features. The combination of experimental evidence and evolutionary information provided a strong rationale for selecting organisms for further study, ultimately leading to the discovery of novel cationic lipid biosynthesis in Enterococcus dispar (Figure 6, Figure S3).

One goal of this study was to find the structural determinants of GlcN-DAG activity by MprF, and we found that hidden unit 2 from our RBM analysis was predictive of this. However, there are notable exceptions such as the E. faecium allele EGTF 00601, which has sequence features placing it in quadrant one yet has GlcN-DAG activity. Therefore, identifying the exact sequence determinants of GlcN-DAG specificity remains to be fully elucidated. One explanation could be our exclusion of the N-terminal flippase domain and the non-cytosolic region of the C-terminal domain in the RBM analysis, where important residue interactions determining specificity could occur and is under further investigation. Additionally, a challenge in experimental validation will be to standardize the method of studying mprF alleles from diverse organisms. When using native organisms, lipids produced may vary during the growth phase of the organisms, the media they are grown in, and whether a membrane stressor is present (i.e. inducible production of specific lipids). For this study we used stationary phase cultures of equal volume and bacteria were grown in their respective standard laboratory media. The use of a heterologous host for expression of mprF alleles is a method of standardization, but may result in not identifying novel lipids that would be found in their native bacteria since the appropriate lipid substrates may not be present [16, 12]. A combination of both approaches (native organisms and heterologous hosts) may be required. Additionally, spike-in controls might be implemented to help us to more quantitatively understand the output specificity of these MprF variants. Ultimately, mutational studies of the residues identified by the RBM will allow for identification of the residues critical for specificity, though likely the residues will depend in part on the primary sequence being altered.

The general method of RBM-guided exploration has applications in enzyme design, for example, in enhancing methods like directed evolution [33]. Concretely, the residues identified through hidden units can be thought of as templates for mutational exploration, greatly reducing the search space for assessing which residues are critical for specificity determination. Enzymes like MprF that produce lipids of modified charge have potential applications in biotechnology and therapeutics, where the design of lipid nanoparticles with different charge and chemical properties of the head group have important physiological implications [34]. Notably, the identification of Lys-Glc2-DAG opens new research avenues, particularly in the fields of antibiotic resistance and host-pathogen interactions, wherein lipid modification and cationic lipids play key roles. With respect to human health, it would be relevant to investigate whether Lys-Glc2-DAG plays a role in enterococcal resistance to the last-line membrane-active antibiotic, daptomycin.

Methods and Materials

Bacterial Strains and Growth Conditions

See Tables S1 and S3 for a full list of bacterial strains used in this study. Streptococcus mitis NCTC12261 (referred to as SM61 here) [35] was grown in Todd Hewitt Broth (THB) at 37°C with 5% CO2. Escherichia coli DH5α [36] and Bacillus subtilis 168 [37] was grown at 37°C with shaking at 225 rpm in lysogeny broth (LB). Staphylococcus aureus RN4220 [38] was grown at 37°C with shaking at 220 rpm in Trypic Soy Broth (TSB). Streptococcus agalactiae COH1 ATCC BAA-1176 [39] and CJB111 [40, 41] were grown in THB at 37 °C. Enterococcus faecium 1,231,410 [42], Enterococcus faecalis T11 [43] and OG1RF [44], Enterococcus gallinarium EG2 [42], Enterococcus casseliflavus EC10 [42], and Enterococcus raffinosus Er676 [45] were grown at 37°C in Brain Heart Infusion (BHI). Enterococcus dispar ATCC 51266 [46] was grown at 30°C in BHI. Exiguobacterium acetylicum UTDF19-27C was grown in BHI at 37 °C. Ligilactobacillus salivarius ATCC 11741 [47], Lacticaseibacillus rhamnosus ATCC 7469 [48], Lactobacillus casei ATCC 393 [49] and Lacticaseibacillus paracasei ATCC 25302 [48] were grown in Lactobacilli MRS Broth at 37°C and 5% CO2. Levilactobacillus brevis ATCC 14869 [50] was grown in Lactobacilli MRS Broth at 30°C, 5% CO2. Bacillus licheniformis ATCC 14580 was grown in nutrient broth at 37°C with shaking at 225 rpm [51]. A transposon mutant of E. faecalis OG1RF from [30] with a mini mariner transposon (EfaMarTn) insertion in OG1RF_10760 (OG1RF_10760::Tn) was grown on BHI supplemented with chloramphenicol at a concentration of 15 μ/mL. The mutant was confirmed to be erythromycin-sensitive. The transposon mutant location was confirmed through Sanger sequencing performed by the Genome Center at the University of Texas at Dallas (Richardson, TX).

Where appropriate for plasmid selection, kanamycin (Sigma-Aldrich) was added. For E. coli, a concentration of 50 μg/mL was used, for S. mitis a concentration of 300 μg/mL, and for E. faecalis a concentration of 500 μg/mL was used.

Routine Molecular Biology Procedures

Genomic DNA was extracted as done previously in [52] and [53]. All PCR reactions used Phusion polymerase (Thermo Fisher Scientific) and Phusion 5x HF buffer (Thermo Fisher Scientific) in a Veriti PCR machine (Applied Biosystems). List of primers used are found in Table S4. Gibson assemblies were completed using a 2x HI-FI Assembly master mix following the manufacturer’s protocol (New England Biolabs). PCR clean-up was done using the GeneJET PCR Purification Kit (Thermo Fisher Scientific) per manufacturer protocol. Plasmid extractions were performed per manufacturer protocol using the GeneJET Plasmid miniprep kit (Thermo Fisher Scientific). All plasmid constructs were confirmed by Sanger sequencing at the Massachusetts General Hospital CCIB DNA Core facility or by Illumina sequence at SeqCenter. DNA concentrations were measured using Nanodrop (Thermo Fisher Scientific) or Quibit 2.0 (Invitrogen by Life Technologies). Optical Density at 600 nm (OD600nm) was measured in a disposable cuvette (Thermo Fisher Scientific) using a spectrophotometer (Thermo Scientific Genesys 30).

Gibson assembly was performed using pABG5 as previously described [11]. Transformation of plasmids into E.coli, S. mitis, and E.faecalis was performed as previously described in [11, 16, 54, 55].

Acidic Bligh-Dyer Lipid Extractions

Bacteria were grown for approximately 16 hours overnight in 15 mL of an appropriate culture medium. After growth, OD600nm measurements were taken, and cells were pelleted at 4,280 x g for 5 minutes at room temperature in a Sorvall RC6+ centrifuge. The supernatant was tipped out, and the cell pellet was washed and resuspended in 1X Phosphate Buffered Saline (PBS). The cells were pelleted again, and all the supernatant was aspirated out. The cell pellet was stored at -80°C until lipid extraction. An acidic Bligh-Dyer lipid extraction was performed as previously reported [16, 12]. Briefly, cell pellets were resuspended in 0.8 mL of 1x Dulbecco’s PBS (Sigma-Aldrich). Cells were transferred to a 9 mL glass tube with a Teflon-lined cap (Pyrex). 1 mL chloroform (MilliporeSigma) and 2 mL methanol (Thermo Fisher Scientific) were added to create the single-phase Bligh-Dyer. Tubes were vortexed every 5 minutes for 20 minutes, then centrifuged at 500 x g for 10 minutes at room temperature. The supernatant was transferred to a new 9 mL glass tube. 100 μL of Hydrochloric acid (Thermo Fisher Scientific) was added followed by 1 mL of chloroform and 0.9 mL of 1x Dulbecco’s PBS to create the two-phase Bligh-Dyer. Tubes were gently mixed and centrifuged at 500 x g for 5 minutes at room temperature. After centrifuging, the bottom layer was extracted to a new tube and dried under nitrogen gas and stored at -80°C prior to lipidomic analysis. Lipid analyses were repeated in biological triplicate aside, from OG1RF_10760::Tn extractions, which were performed once.

Analysis of lysine lipids by an amino column-based Liquid Chromatography-Electrospray Ionization-Mass Spectrometry (LC-ESI MS): A new method

An amino column-based normal phase LC-ESI MS was performed using an Agilent 1200 Quaternary LC system coupled to a high-resolution TripleTOF5600 mass spectrometer (Sciex, Framingham, MA). A Unison UK-Amino column (3 μm, 25 cm × 2 mm) (Imtakt USA, Portland, OR) was used. Mobile phase A consisted of chloroform/methanol/aqueous ammonium hydroxide (800:195:5, v/v/v). Mobile phase B consisted of chloroform/methanol/water/ aqueous ammonium hydroxide (600:340:50:5, v/v/v/v). Mobile phase C consisted of chloroform/methanol/water/aqueous ammonium hydroxide (450:450:95:5, v/v/v/v). The elution program consisted of the following: 100% mobile phase A was held isocratically for 2 min and then linearly increased to 100% mobile phase B over 8 min and held at 100% B for 5 min. The LC gradient was then changed to 100% mobile phase C over 1 min and held at 100% C for 3 min, and finally returned to 100% A over 0.5 min and held at 100% A for 3 min. The total LC flow rate was 300 μl/min. The MS settings were as follows: Ion spray voltage (IS) = 5000 V, Curtain gas (CUR) = 20 psi, Ion source gas 1 (GS1) = 20 psi, De-clustering potential (DP) = 50 V, and Focusing Potential (FP) = 150 V. Nitrogen was used as the collision gas for MS/MS experiments. Data acquisition and analysis were performed using Analyst TF1.5 software (Sciex, Framingham, MA).

Generation of mprF expression plasmids

The S. downei mprF nucleotide sequence (Locus tag HMPREF9176 RS03810) and S. ferus mprF nucleotide sequence (Locus tag A3GY RS0106165) were used to design synthetic geneblocks (GeneWiz from Azenta Life Sciences). A 20 nucleotide 5’ extension and a 20 nucleotide 3’ extension complementary to the pABG5 plasmid were added to enable Gibson assembly of the inserts with pABG5. Sequences for the geneblocks can be found in Table S5. The S. salivarius mprF (Locus tag SSAL8618 04345) was amplified from Streptococcus salivarius ATCC 7073. A 20 nucleotide 5’ extension and a 20 nucleotide 3’ extension complementary to the pABG5 plasmid was added to the insert. The same protocol was used for generation of pSobrinus, pEFM-prF1, pEFMprF2, and pOGMprF2 with slight modifications. S. sobrinus mprF (Locus tag DLJ52 05040) was amplified from Streptococcus sobrinus ATCC 27352. E. faecium mprF1 (Locus tag EFTG_00601) and E. faecium mprF2 (Locus tag EFTG_02430) were amplified from Enterococcus faecium 1, 231, 410. E. faecalis mprF2 (Locus tag OG1RF_10760) was amplified from E. faecalis OG1RF.

Boltzmann Machines and Restricted Boltzmann Machines for Protein Sequences

Formulation of models

The Boltzmann Machine, as described in [18], is defined as: which is a probability distribution defined on the pairwise interactions (the couplings eij) between the positions (vn) in an input sequence (v) and a local field term hi. The pairwise coupling matrix is akin to a covariance matrix, and the local fields are similar to single site frequency measures.

The joint probability distribution of the Restricted Boltzmann Machine as described in [14] is as follows: Here, the vector v again represents the input sequence data, g(vi) is a local field which controls the conditional probability of the input data,𝒰μ is the hidden unit potential/activation, and the terms hi and wiμ(v) couple the input variables with the hidden unit variables. In this way, interactions between residues are mediated by a relatively smaller set of hidden units which do not interact directly with each other, leading to a bipartite structure as opposed to the Boltzmann machine’s fully connected pair based formulation.

Particularly important is the potential function 𝒰, which in this work is the dRELU function defined as: The four parameters γμ,±, θμ,± are learned through the inference procedure, and allow the positive and negative components of the hidden units wμ to be separately gated, which allows learning of bimodal distributions for the activations hμ. Another key feature is the sparsity regularization on the weights applied during model inference: where q is the number of amino acids plus a gap character (21) and the hyperparameter can be increased to induce sparsity and lower weight magnitude. Equations 4 and 5 together guide the representation of the hidden units to bimodal outputs with interpretable features.

Dataset acquisition and preprocessing

Multiple sequence alignment used for model training was the DUF2156 domain acquired from Pfam [25] (in later revisions renamed LPG synthase C). Starting with this MSA, it was processed to include only amino acid characters and gaps, excluding sequences with ambiguous or non-standard characters. Additionally, sequences with contiguous strings of gaps greater than 20% of the full sequence’s length were removed. This left 11,507 sequences with a length of 298. After the data cleaning process described in Figure 4, a total of 7,865 sequences were used to train our final model.

Model training

The model previously described [14] was utilized, specifically the Python 2.7 version freely available on Github (https://github.com/jertubiana/ProteinMotifRBM). Extensive testing of the combinations of the number of hidden units, L2/L1 regularization, and learning rate were performed. Models were trained in triplicate with different random seeds, as the training procedure converges to slightly different weight values depending on their initial random configuration. Training was considered successful when the model consistently produced similar sets of weights across different random seeds, and the weights had the previously mentioned sparsity and magnitude indicating compositional representation. The final model used in this work was trained for 2000 epochs with a learning rate of 0.1, a learning rate decay of 0.33, L2/L1 parameter set to 1.0, and 300 hidden units.

Model Weight selection

To find hidden units which were relevant to function we computed the L1 norm of each hidden unit individually, then ranked them. The hidden units with the largest magnitude were looked at first, and these were typically the sparsest of all hidden units. Hidden units were chosen which involved sequence coordinates implicated in our function of interest. Specifically, locations identified through Autodock [26] where the lipid was likely to interact, and a small radius around this region to select a small set of coordinates. We chose the only hidden units which had both overlap with multiple residues in this chosen radius and predictive power (separation) for the three examples we had to start with.

Software Accessibility

DUF2156 domain sequences were retrieved from Pfam33 [25]. Raw sequence datasets are available as Supplementary Material.